---
title: "R Markdown"
author: "Thomas Holland (th675)"
date: "December 22, 2022"
output: 
    rmdformats::html_clean:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        code_folding: show
        self_contained: true
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# disable warnings
options(warn=-1)

# libraries used can be installed using the following commands
# install.packages("tidyverse")
# install.packages("rmarkdown")
# install.packages("rmdformats") # for custom html output, using the html_clean format

# load libraries
library(tidyverse)
library(knitr)
```

Click the images to view larger versions.

# Inital data import

## Import data

```{r}
# load data from data.csv using read_csv from the tidyverse package
data <- read_csv("data.csv")

kable(head(data))
```

## Data cleaning

```{r}
# print number of rows with missing values
sum(is.na(data))
```

## Log transformation of size data

```{r}
# add new column to data with natural log transformed length

data <- data %>% 
    mutate(log_length = log(length))

kable(head(data))

```

# Basic examination of data

## Box plot of data

```{r}
# box plot of length, grouped by sex from the dataset data, using the ggplot2 package. 

ggplot(data, aes(length, sex)) + 
    geom_boxplot() + 
    geom_jitter(width = 0, height = 0.2, alpha = 0.2) +  # added jitter so overlapping points are visible
    theme_bw() +
    labs(title="Box plot of lengths grouped by the sex of the organism", x="Length (mm)" , y="Sex")
```

```{r}
# box plot of lengths, grouped by sex from the dataset data, using the ggplot2 package. Add x scale every 10 mm

ggplot(data, aes(length, sex)) + 
    geom_boxplot() + 
    geom_jitter(width = 0, height = 0.2, alpha = 0.2) +  # added jitter so overlapping points are visible
    theme_bw() +
    coord_trans(x = "log") + # transform the x axis to natural  log scale
    scale_x_continuous(breaks = seq(0, 100, 10)) + # add more values to the x axis
    labs(title="Box plot of log adjusted lengths grouped by the sex of the organism", x="Length (mm), log scale" , y="Sex")
```

# Statistical tests

This section contains the results of the statistical tests carried out on the data. The analysis of these results and comments are in the analysis section.

## Binomial test

Carry out and report the results of a binomial test to evaluate whether the sex ratio of the sample deviates from 1:1.

```{r}
binomial <- binom.test(table(data$sex), p = 0.5, alternative = "two.sided", conf.level = 0.95)

binomial
```

## T-test

Carry out and report the results of a t-test to evaluate whether females and males differ in size.

```{r}
lm.t_test <- lm(data$length ~ data$sex)
lm.t_test
summary(lm.t_test)

t.test(data$length ~ data$sex, alternative = "two.sided", conf.level = 0.95, var.equal = TRUE)
```

### Diagnostic plots for t-test

Use histogram for diagnostic plots for the t-test.

```{r}
ggplot(data = data, aes(x = length)) + 
    geom_histogram(aes(y = ..density..), bins = 30, fill = "grey") + 
    geom_density(alpha = 0.2, fill = "grey") + 
    facet_wrap(~sex, ncol = 1) + 
    theme_bw() +
    labs(title="Histogram of lengths grouped by sex", x="Length (mm)" , y="Density")
```

Also use the built-in diagnostic plots for the t-test.

```{r fig.dim = c(10, 15)}
par(mfrow = c(3,2))
hist(resid(lm.t_test), main = "Residuals", xlab = "Residuals", ylab = "Frequency") 
plot(lm.t_test)
```

## Mann-Whitney U test

Non-parametric alternative to the t-test.

```{r}
mannwhitney <- wilcox.test(data$length ~ data$sex, alternative = "two.sided", conf.level = 0.95)

mannwhitney
```

The diagnostic plots for the t-test can also be used for the Mann-Whitney U test.

## T-test on log transformed data

```{r}
lm.log_t_test <- lm(log(data$length) ~ data$sex)
lm.log_t_test

summary(lm.log_t_test)

t.test(log(data$length) ~ data$sex, alternative = "two.sided", conf.level = 0.95, var.equal = TRUE)

t.test(log(data$length) ~ data$sex, alternative = "two.sided", conf.level = 0.95, var.equal = FALSE)
```

### Diagnostic plots for t-test on log transformed data

```{r}
ggplot(data = data, aes(x = length)) + 
    geom_histogram(aes(y = ..density..), bins = 30, fill = "grey") + 
    geom_density(alpha = 0.2, fill = "grey") + 
    facet_wrap(~sex, ncol = 1) + 
    theme_bw() +
    coord_trans(x = "log") + # transform the x axis to natural  log scale
    scale_x_continuous(breaks = seq(0, 100, 10)) + # add more values to the x axis
    labs(title="Histogram of lengths grouped by sex", x="Length (mm), log scale" , y="Density")
```


Utilising the built-in diagnostic plots for the t-test.

```{r fig.dim = c(10, 15)}
par(mfrow = c(3,2))
hist(resid(lm.log_t_test), main = "Residuals", xlab = "Residuals", ylab = "Frequency")
plot(lm.log_t_test)
```

# Analysis

## Binomial test

The binomial test confirmed that there was no significant deviation from the sex ratio displayed in the data because the p-value of `r binomial$p.value` is greater than the significance level of 0.05.

## T-test

The effect of sex on the length of the organism was tested, and it was found that males were `r round(lm.t_test$coefficients[2], 2)` mm longer than the females and that the females were found to be `r round(lm.t_test$coefficients[1], 2)` mm long on average (F= `r summary(lm.t_test)$fstatistic[1]`, p-value of `r summary(lm.t_test)$fstatistic[4]`).

However, the diagnostic plots for the linear model generated from this is not ideal. Firstly, the histogram plot shows an unequal variance, as the male group has a much larger variance. Also, the histogram plot shows that there is a non-normal distribution with the male having a possible positive skew.

This is also seen with the residuals vs fitted diagnostic plot that has a much smaller spread around the female group (the lower fitted value). Also in the normal Q-Q plot, the datapoints stray quite far from the 1:1 expectation line, especially at either end suggesting that a transformation may be necessary.

This is confirmed as the Mann-Whitney U test had a non-significant result (p-value of `r mannwhitney$p.value`), confirming that even without parametric assumptions the data is not significantly different, and so a transformation may be necessary.

However, based on the linear model generated using the log-transformations, there is no significant predictive value of sex in the length of the fish (p-value of `r t.test(log(data$length) ~ data$sex, alternative = "two.sided", conf.level = 0.95, var.equal = TRUE)$p.value`, equal variance), and even when testing without the assumption of equal variance the predictive value is still not significant (p-value of `r t.test(log(data$length) ~ data$sex, alternative = "two.sided", conf.level = 0.95, var.equal = FALSE)$p.value`).

*Please note that there is "hidden" code within this that is used to access the p-values by using the following code as an example: `` `r "\u0060r mannwhitney$p.value\u0060"` ``*